Handling String-Number Conversion in String Constraint Solving

### 2019-11-06 draft.tex
evaluation section: initial
polish the definition of string constraints

\usepackage{xcolor}
\usepackage{amssymb} 
\usepackage{tikz}
\usepackage[autostyle]{csquotes}
\usepackage{listings}
\synctex=1

@@

\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\lh}[1]{{\color{orange}Lukas: #1}}
\newcommand{\sti}[1]{\mbox{\textbf{toInt}($#1$)}}
\newcommand{\its}[1]{\mbox{\textbf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{\mbox{$V^1_{\Sigma_\epsilon}$}}
\newcommand{\cvartwo}{\mbox{$V^2_{\Sigma_\epsilon}$}}
\newcommand{\modelof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}
\newcommand{\parikhof}[1]{\mathit{Parikh}{(#1)}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Preliminaries} \label{section:preliminary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%We use $\mathbb{N}$ and $\mathbb{Z}$ to denote the sets of natural numbers and 
%integers. For a set $A$, we use $|A|$ to denote its size. For a string $w$, we 
%use $|w|$ to denote its length and $|w|_a$ to denote the number of occurrences of $a$ in $w$. We $\epsilon$ to denote an empty string and use $w_1\cdot w_2$ to denote the concatenation of strings $w_1$ and $w_2$. Let $S$ be a finite set of symbols. We use $S^+$ to denote the set of string over $S$ and $S^* = S^+\cup \{\epsilon\}$. We define $S_\epsilon =S\cup\{\epsilon\}$. A language 
%$L$ over $S$ is a set of strings in $S^*$. 
%
%A \emph{finite automaton} is a tuple $(Q,T,\Sigma,q_0,q_m)$, where $Q$ is the set of states, $T\subseteq Q\times (\Sigma\cup \{\epsilon\} )\times Q $ is the set of transition relation, $\Sigma$ is the alphabet, $q_0$ is the initial state, and $q_m$ is the final state. The semantics of finite automaton is defined in the standard manner. We use $L(A)$ to denote the regular language, i.e., set of accepted strings, of the automaton $A$.

%Given a finite automaton $A$ over $\{a_1,a_2,\ldots,a_n\}$, the Parikh image of $L$ is the set $\{(|w|_{a_1},\ldots, |w|_{a_n}) \mid w \in  L(A) \}$, i.e., the number of occurrences of each symbol $\{a_1,a_2,\ldots,a_n\}$ of words in $L(A)$. Moreover, one can construct from $A$ in linear time a Presburger formula $\phi_A(v_1, \ldots,v_n)$ such that $\phi(c_1, \ldots,c_n) \iff (c_1, \ldots,c_n)$ in the Parikh image of $L$~\cite{SeidlSMH04}.

We use $\mathbb{N}$ and $\mathbb{Z}$ to denote the sets of natural numbers (including 0) and 
integers. For a set $A$, we use $|A|$ to denote its size. 
An \emph{alphabet} is a finite set $\Sigma$ of \emph{characters} and a \emph{string} over $\Sigma$ is a sequence $w = a_1\ldots a_n$ of characters from $\Sigma$, with $\epsilon$ denoting the \emph{empty string}. 
We use $w_1\cdot w_2$ to denote the \emph{concatenation} of strings $w_1$ and $w_2$.
$\Sigma^*$ is the set of all strings over $\Sigma$, $\Sigma^+ = \Sigma^*\setminus \{\epsilon\}$, and 
A \emph{language} over $\Sigma$ is a subset $L$ of $\Sigma^*$.  
%
We use $|w|$ to denote the lenght of $w$ and $|w|_a$ to denote the number of occurrences of the character $a\in \Sigma$ in $w$.  
The \emph{Parikh image} of $w$ is the assignment 
%$\parikh(w) = \{a_1 \mapsto |w|_{a_1},\ldots, a_n \mapsto |w|_{a_n}\}$ that maps every symbol in $w$ to the number of its occurences in $w$. 
$\parikhof w$ that maps every character $a\in\Sigma$ to the number of its occurences in $w$, i.e., $\parikhof w (a) = |w|_a$. 
The Parikh image of a language $L$ is the set of Parikh images $\parikhof L = \{\parikhof w \mid w \in  L\}$.

A \emph{finite automaton} is a tuple $(Q,T,\Sigma,q_0,q_m)$, where $Q$ is the set of \emph{states}, $T\subseteq Q\times (\Sigma\cup \{\epsilon\} )\times Q $ is the set of \emph{transitions}, $\Sigma$ is the alphabet, $q_0$ is the \emph{initial state}, and $q_m$ is the \emph{final state}. 
%The semantics of finite automaton is defined in the standard manner. 
We use $L(A)$ to denote the language of the automaton $A$, i.e., the set of accepted strings, defined in the standard manner.
\lh{If we later start technical talking about tings like runs then we should include this definition.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Symbolic Flat Automata} \label{section:sfa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\st}[2]{q_{#1}^{#2}}
\newcommand{\sym}[2]{v_{#1}^{#2}}
%We use $\cvar$ to denote a finite set of \emph{character variables}  ranging over $\Sigma_\epsilon$. 
%Sometimes we use index to denote different set of variables over the same range, e.g., $\cvarone$ and $\cvartwo$. 
%We use $v$ and its variants for character variables. We call a string over $\cvar$ (character variables) a \emph{symbolic string}.
%An \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$. For a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

We introduce \emph{symbolic flat automata} (SFA) that will be used to define patterns 
%for the string variable domain restriction. 
used by the underapproximation module to restrict the domain of string variables. 
Namely, an SFA is a finite automaton $A = (Q,T,\cvar,\st{0}{0},\st{m}{0})$ operating over the alphabet $\cvar$ of \emph{character variables}, 
and satisfying the following structural constraints:
\begin{enumerate}
	\item The final state $\st{m}{0}$ is reached from  the initial state $\st{0}{0}$ through a stright path of $m-1$ transitions $(\st{i}{0},v_i,\st{i+1}{0}) \in T$, $\st{i}{0} \in Q$ for $i\in[0,m-1]$. %, with pairwise distiand states $\st{i}{0} \in Q$ for $i\in[0,m-1]$.
	\item  
Each state $\st{i}{0}$ is the origin of a unique simple cycle of the length $l_i\in\mathbb{N}$, consisting of states $\st{i}{j} \in Q$ and transitions $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$ for $j\in [1,l_i]$. 
Notice that case when $l_i = 0$ is also admissible and means that there is no cycle on $q_i$.
%Assume that $\st{i}{0}$ is included in a cycle with $l_i$ states (includes $\st{i}{0}$). 
%For each $j\in [1,l_i]$, we have $\st{i}{j} \in Q$ and $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$. 
%(where $l_i = 0$ means that there is no cycle on $q_i$)
	\item Each variable in $\cvar$ appears on at most one transition (but $\epsilon$ transition can appear multiple times). 
\end{enumerate} 
%Such automata are recognizers of symbolic strings, i.e., strings over $\cvar$.
Symbolic automata accept strings over $\cvar$, called \emph{symbolic strings}. 
Each symbolic string may be interpeted as a string over $\Sigma$, given an assignment of characters to its character variables.  
%
Formally, an \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$ and for a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

@@

%A finite automaton $D=(Q,T',\Sigma,q_0,q_m)$ is an \emph{instance} of a SFA $S=(Q,T,\cvar,q_0,q_m)$, if there exists an interpretation $I: \cvar \cup \{\epsilon\} \rightarrow \Sigma_\epsilon$ satisfying $I(\epsilon)=\epsilon$ and $T'=\{(q,I(c),q')\mid (q,c,q')\in T\}$. We write $I(S)$ to denote the instance of $S$ wrt. $I$. The language of $S$ is defined as $L(S)= \cup\{L(D) \mid D \mbox{ is an instance of } S\}$. Notice that $L(S)$ is regular because every SFA has only finitely many instance and the finite union of regular languages are also regular.

The purpose of the structural restrictions of SFA is to ensure that the words of $L(A)$ are uniquely determined by their Parikh images, as formalised by the following lemma.

\begin{lemma}
%Distinct words from the language of an SFA must have distinct Parikh images.  
For an SFA $A$ and $x,y\in L(A)$ if $\parikhof x = \parikhof y$ then $x = y$. 
%Distict words from the language of an SFA have distinct Parikh images.  
\end{lemma}
Indeed, notice that each accepted symbolic string $x$ is determined by the numbers of repetitions of individual cycles in the accepting run of $A$ over $x$.
Since every variable appears on at most one transition, then the $\parikhof x$ value of all variables appearing within the same cycle is the same, and it is equal to the number of repetitions of that cycle in the accepting run. 
%
The accepting run on $x$ (and so $x$ itself) can thus be reconstructed from $\parikhof x$.

For example, in the example of Figure~\ref{fig:sfa_def}, 
from $\parikhof x(v_0) = \parikhof x(v_1) = \cdots = \parikhof x(v_{m-1})=1$ and $\parikhof x(v^0_1)=\parikhof x(v^1_1)=\parikhof x(v^2_1)=2$, then $x=v_0(v^0_1v^1_1v^2_1)^2v_1\cdots v_{m-1}$. 
\todo{add Presburger constraints to allow more expressive power.}

\lh{so far not sure about how and where to present this:}
\begin{lemma}
	Symbolic flat automata are closed under concatenation.
\end{lemma}

>>>>

\usepackage{xcolor}
\usepackage{amssymb} 
\usepackage{tikz}
\usepackage[autostyle]{csquotes}
\usepackage{listings}

@@

\newcommand{\todo}[1]{{\color{blue}TODO: #1}}
\newcommand{\sti}[1]{\mbox{\textsf{toInt}($#1$)}}
\newcommand{\its}[1]{\mbox{\textsf{toStr}($#1$)}}
\newcommand{\varn}{\mbox{$\mathbb{V}_{\mathbb{Z}}$}}
\newcommand{\vars}{\mbox{$\mathbb{V}_{\Sigma^*}$}}
\newcommand{\cvar}{\mbox{$V_{\Sigma_\epsilon}$}}
\newcommand{\cvarone}{\mbox{$V^1_{\Sigma_\epsilon}$}}
\newcommand{\cvartwo}{\mbox{$V^2_{\Sigma_\epsilon}$}}
\newcommand{\modelof}[1]{[\![#1]\!]}
\newcommand{\true}{\mbox{$\mathsf{true}$}}
\newcommand{\false}{\mbox{$\mathsf{false}$}}
\newcommand{\enc}[1]{[\![#1]\!]}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Preliminary} \label{section:preliminary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We use $\mathbb{N}$ and $\mathbb{Z}$ to denote the sets of natural numbers and 
integers. For a set $A$, we use $|A|$ to denote its size. For a string $w$, we 
use $|w|$ to denote its length and $|w|_a$ to denote the number of occurrences of $a$ in $w$. We $\epsilon$ to denote an empty string and use $w_1\cdot w_2$ to denote the concatenation of strings $w_1$ and $w_2$. Let $S$ be a finite set of symbols. We use $S^+$ to denote the set of string over $S$ and $S^* = S^+\cup \{\epsilon\}$. We define $S_\epsilon =S\cup\{\epsilon\}$. A language 
$L$ over $S$ is a set of strings in $S^*$.

A \emph{finite automaton} is a tuple $(Q,T,\Sigma,q_0,q_m)$, where $Q$ is the set of states, $T\subseteq Q\times (\Sigma\cup \{\epsilon\} )\times Q $ is the set of transition relation, $\Sigma$ is the alphabet, $q_0$ is the initial state, and $q_m$ is the final state. The semantics of finite automaton is defined in the standard manner. We use $L(A)$ to denote the regular language, i.e., set of accepted strings, of the automaton $A$.

Given a finite automaton $A$ over $\{a_1,a_2,\ldots,a_n\}$, the Parikh image of $L$ is the set $\{(|w|_{a_1},\ldots, |w|_{a_n}) \mid w \in  L(A) \}$, i.e., the number of occurrences of each symbol $\{a_1,a_2,\ldots,a_n\}$ of words in $L(A)$. Moreover, one can construct from $A$ in linear time a Presburger formula $\phi_A(v_1, \ldots,v_n)$ such that $\phi(c_1, \ldots,c_n) \iff (c_1, \ldots,c_n)$ in the Parikh image of $L$~\cite{SeidlSMH04}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Symbolic Flat Automata} \label{section:sfa}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\st}[2]{q_{#1}^{#2}}
\newcommand{\sym}[2]{v_{#1}^{#2}}
We use $\cvar$ to denote a finite set of \emph{character variables}  ranging over $\Sigma_\epsilon$. Sometimes we use index to denote different set of variables over the same range, e.g., $\cvarone$ and $\cvartwo$. We use $v$ and its variants for character variables. We call a string over $\cvar$ (character variables) a \emph{symbolic string}.
An \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$. For a symbolic string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.

We introduce \emph{symbolic flat automata} (SFA) $(Q,T,\cvar,\st{0}{0},\st{m}{0})$ that will be used to define patterns for the string variable domain restriction. SFA is a finite automaton operating over the alphabet $\cvar$ in addition with the following structural constraints.
\begin{enumerate}
	\item the initial state $\st{0}{0}$ and final state $\st{m}{0}$ are connected by a straight-line structure, i.e., $(\st{i}{0},v_i,\st{i+1}{0}) \in T$ and states $\st{i}{0} \in Q$ for $i\in[0,m-1]$.
	\item  each state $\st{i}{0}$ may be included to an unique simple cycle. Assume that $\st{i}{0}$ is included in a cycle with $l_i$ states (includes $\st{i}{0}$), then we have $\st{i}{j} \in Q$ and $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$, for $j\in [1,l_i]$. 
	\item each variable in $\cvar$ appears in at most one transition (but $\epsilon$ transition can appear multiple times). 
\end{enumerate} 
Such automata are recognizers of symbolic strings, i.e., strings over $\cvar$.

@@

A finite automaton $D=(Q,T',\Sigma,q_0,q_m)$ is an \emph{instance} of a SFA $S=(Q,T,\cvar,q_0,q_m)$, if there exists an interpretation $I: \cvar \cup \{\epsilon\} \rightarrow \Sigma_\epsilon$ satisfying $I(\epsilon)=\epsilon$ and $T'=\{(q,I(c),q')\mid (q,c,q')\in T\}$. We write $I(S)$ to denote the instance of $S$ wrt. $I$. The language of $S$ is defined as $L(S)= \cup\{L(D) \mid D \mbox{ is an instance of } S\}$. Notice that $L(S)$ is regular because every SFA has only finitely many instance and the finite union of regular languages are also regular.

For such kind of automata, the number of occurrence of each symbol in $\cvar$ uniquely characterize an accepting symbolic string. For example, in the example of Figure~\ref{fig:sfa_def}, if the string $v_0(v^0_1v^1_1v^2_1)^2v_1v_{m-1}$ is accepted, then all other strings $w$ with $|w|_{v_0}=|w|_{v_1}=\ldots=|w|_{v_{m-1}}=1$ and also $|w|_{v^0_1}=|w|_{v^1_1}=|w|_{v^2_1}=2$ are not accepted. Moreover, from $|w|_{v_0}=|w|_{v_1}=\ldots=|w|_{v_{m-1}}=1$ and $|w|_{v^0_1}=|w|_{v^1_1}=|w|_{v^2_1}=2$, we can derive $w=v_0(v^0_1v^1_1v^2_1)^2v_1v_{m-1}$ by exploring the automata structure. \todo{add Presburger constraints to allow more expressive power.}

\begin{lemma}
	Symbolic flat automata are closed under concatenation.
\end{lemma}

\todo{Mention NDD when talk about our practical bounded encoding for str2int}
